---
title: "Model selection for vascular plant cover"
author: "Claire Powers"
date: '2023-05-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Packages and data
```{r start}
rm(list=ls())

### Packages
library(tidyverse)
library(lme4)
library(MuMIn)
library(rje)
library(performance)
library(sjPlot)
library(corrplot)

### Read in and clean data

dats.orig=read_csv('./AnalysisData/vascplantcover_final.csv') %>% 
  filter(!is.na(start.plantCover)) %>% 
  mutate(PSaspect=paste(park,summit,aspect,sep="."),
         pchangecover = (change.cover/start.plantCover)*100) %>% 
  relocate(c(change.cover,pchangecover),.after = end.plantCover) %>% 
    arrange(-pchangecover) %>% 
  filter(pchangecover<100) %>% 
  mutate(park = factor(park,levels=c("GNP","YNP","RMN","GSD","PEC")),
         tr.num = str_sub(transition,4,4))

dats = dats.orig %>% 
  select(-contains(c("rock","scree"))) %>% 
  na.omit()

### Create another dataframe where numeric predictor variables are scaled ###
dats.scaled = dats # Get names columns to standardize
scale_cols = dats %>% select(start.plantCover,climPC1,climPC2,sitelat,contains(c(".tmean",".aet",".pet",".ppt",".snowdays",".gdd"))) %>% names()
dats.scaled[scale_cols]=lapply(dats.scaled[scale_cols], function(x) scale(x))

rm(scale_cols)
```

##### Looking at vascular plant cover data
```{r}
names(dats.orig)
ggplot(dats.orig,aes(park,y=change.cover)) +
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  theme_classic(base_size =18)

ggplot(dats.orig,aes(x=mean.snowdays,y=change.cover)) +
  geom_hline(yintercept = 0)+
  geom_point()+
  facet_wrap(~park)

ggplot(dats.orig,aes(x=start.plantCover,y=end.plantCover,color=tr.num))+
  geom_point()+
  geom_abline(slope=1)+
  facet_wrap(~park)+
  theme_classic()

```


Vascular plant cover models have very few predictor variables, so model selection is much simpler
- Fixed effects are start cover, climate variables, site latitude, and park -- here
- Random effect options are PSaspect and transition
```{r REstructure}

### Compare RE options with a fully saturated FE structure

# PS aspect random slop
mod1 = lmer(end.plantCover~start.plantCover*(climPC1+climPC2+sitelat) + park + (1|PSaspect),
                  na.action = "na.fail",data = dats.scaled,REML = TRUE)

# transition as random slope
mod2 = lmer(end.plantCover~start.plantCover*(climPC1+climPC2+sitelat) + park + (1|transition),
                  na.action = "na.fail",data = dats.scaled,REML = TRUE)

# both as random slopes
mod3 = lmer(end.plantCover~start.plantCover*(climPC1+climPC2+sitelat) + park + (1|PSaspect) + (1|transition),
                  na.action = "na.fail",data = dats.scaled,REML = TRUE)

## Same as above but without park ##
mod4 = lmer(end.plantCover~start.plantCover*(climPC1+climPC2+sitelat) + (1|PSaspect),
                  na.action = "na.fail",data = dats.scaled,REML = TRUE)

# transition as random slope
mod5 = lmer(end.plantCover~start.plantCover*(climPC1+climPC2+sitelat) + (1|transition),
                  na.action = "na.fail",data = dats.scaled,REML = TRUE)

# both as random slopes
mod6 = lmer(end.plantCover~start.plantCover*(climPC1+climPC2+sitelat) + (1|PSaspect) + (1|transition),
                  na.action = "na.fail",data = dats.scaled,REML = TRUE)

# Compare AICc values
mod.compare = AICc(mod1,mod2,mod3,mod4,mod5,mod6) %>% mutate(delta = AICc-min(AICc)) %>% arrange(delta)

##### Mod 2 -- transition as a random slope -- is best. Now refit with ML and dredge
mod2.ML = lmer(end.plantCover~start.plantCover*(climPC1+climPC2+sitelat) + park + (1|transition),
                  na.action = "na.fail",data = dats.scaled,REML = F)

## Dredge the model that was refit with maximum likelihood
mod2.dredge = dredge(mod2.ML,beta = c("none"),fixed = c('start.plantCover'),trace=2)
#####

### The best model here does not include and climPC variables. Do we want to just fix them? The second best model does include clim PC2

##### Pull best model from dredge #####
best.PC.mod = get.models(mod2.dredge,subset = 1)[[1]]
best.PC.mod@call

# end.plantCover ~ start.plantCover + climPC1 + climPC1:start.plantCover + park + (1 | transition) 
# rm(as.vector(rownames(mod.compare)[2:7]))
rm(mod1,mod2,mod3,mod4,mod5,mod6,mod.compare,mod2.dredge,mod2.ML)
```


Sub in climate vars for climPC1

3. Take that model with the optimal FE and RE structure, remove random slopes and find best explicit climate variables in nested loops
```{r best.clim.vars, warning=FALSE}

# Create climate variable variable groups
val = c("min","max","mean")
tmean.grp = append(paste0(val,".tmean"),NA)
snow.gdd.grp = append(c(paste0(val,".gdd"),paste0(val,".snowdays")),NA)
ppt.grp = append(paste0(val,".gs.ppt"),NA)
cwd.grp =  append(c(paste0(val,".aet"),paste0(val,".pet"),paste0(val,".wy.ppt")),NA)
rm(val)

# to see number of combinations
clim.combos.df = expand_grid(tmean.grp,snow.gdd.grp,ppt.grp,cwd.grp) %>% filter(rowSums(is.na(.))!=ncol(.))

# Use the best global model from above to sub in explicit climate variables
# mod2 = lmer(end.plantCover~start.plantCover*(climPC1+climPC2+sitelat) + park + (1|transition),
#                   na.action = "na.fail",data = dats.scaled,REML = TRUE)

climVar.generic.mod = "end.plantCover~start.plantCover*(tmean.var + cwd.var + ppt.var + snow.gdd.var + sitelat) + park + (1|transition)"

# In each iteration, a single climate variable from each group is subbed in for the generic climate variables above (i.e., tmean.var is swapped out for mean.tmean). That model is then fit with ML (not REML) and the AIC value is stored in the output table for comparison once the loops are finished running

# Store results
clim.mod.AICs = NULL

for(i in 1:nrow(clim.combos.df)){
  
  tmean.i = clim.combos.df$tmean.grp[i]
  cwd.i = clim.combos.df$cwd.grp[i]
  ppt.i = clim.combos.df$ppt.grp[i]
  snow.gdd.i = clim.combos.df$snow.gdd.grp[i]

  # If all clim vals are NA skip to next loop
  if(is.na(tmean.i)&is.na(cwd.i)&is.na(ppt.i)&is.na(snow.gdd.i)) next
  
  clim.mod.i = climVar.generic.mod
  
  # either remove tmean.var or replace with specific tmean variable name. Repeat the same process for all other clim variable groups
  if(is.na(tmean.i)) {clim.mod.i = gsub(pattern = "tmean.var + ",replacement = "", x = clim.mod.i, fixed = T)
  }else{clim.mod.i = gsub(pattern = "tmean.var",replacement = tmean.i,x = clim.mod.i)}
  
  if(is.na(cwd.i)) {clim.mod.i = gsub(pattern = "cwd.var + ",replacement = "", x = clim.mod.i, fixed = T)
  }else{clim.mod.i = gsub(pattern = "cwd.var",replacement = cwd.i,x = clim.mod.i)}
          
  if(is.na(ppt.i)) {clim.mod.i = gsub(pattern = "ppt.var + ",replacement = "", x = clim.mod.i, fixed = T)
  }else{clim.mod.i = gsub(pattern = "ppt.var",replacement = ppt.i,x = clim.mod.i)}
  
  if(is.na(snow.gdd.i)) {clim.mod.i = gsub(pattern = " + snow.gdd.var",replacement = "", x = clim.mod.i, fixed = T)
  }else{clim.mod.i = gsub(pattern = "snow.gdd.var",replacement = snow.gdd.i,x = clim.mod.i)}
  
  out_df = data.frame(tmean.var = tmean.i,
              snow.gdd.var = snow.gdd.i,
              ppt.var = ppt.i,
              cwd.var = cwd.i,
              mod.str = clim.mod.i)
  
  # Fit in LMER
  mod.fit.i = lmer(formula = clim.mod.i, data = dats.scaled, REML = FALSE, na.action = "na.fail")

  out_df$AICc=AICc(mod.fit.i)
  out_df$condR2 = r2(mod.fit.i)[[1]]
  out_df$margR2 = r.squaredGLMM(mod.fit.i)[[2]]

  clim.mod.AICs = rbind(clim.mod.AICs,out_df)
  rm(out_df,clim.mod.i)

}

## add a delta AIC column
clim.mod.AICs$delta = clim.mod.AICs$AICc-min(clim.mod.AICs$AICc)

## Pull out model string for model with lowest AIC
best.ExplClimVar.mod.str <- clim.mod.AICs$mod.str[clim.mod.AICs$delta==0]

## Fit that model again and put in global env
best.ExplClimVar.mod = lmer(best.ExplClimVar.mod.str, data = dats.scaled, REML = FALSE, na.action = "na.fail")

## Dredge that model
dredge.climVar.mod = dredge(best.ExplClimVar.mod,beta = c("none"),fixed = c('start.plantCover'),trace=2)

## Pull out best model from dredge and refit with REML for parameter estimates
best.climVar.mod = get.models(dredge.climVar.mod,subset=1)[[1]]
best.climVar.mod@call

final.climVar.mod = lmer(formula = end.plantCover ~ max.snowdays + mean.gs.ppt + min.wy.ppt + park + start.plantCover + max.snowdays:start.plantCover + mean.gs.ppt:start.plantCover + min.wy.ppt:start.plantCover + (1 | transition), data = dats.scaled, REML = FALSE, na.action = "na.fail")
summary(final.climVar.mod)

```

### Best vascular plant cover models ###
climate pca
```{r}

tab_model(final.climVar.mod)
plot_model(final.climVar.mod)

# No transition random effects are significantly different from zero
plot_model(final.climVar.mod,type="re")+
  geom_vline(xintercept = 0)+
  theme_bw()
```


