---
title: "Climate data interpolation"
author: "Claire Powers"
date: "1/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Clear env
```{r clearenv}
rm(list=ls())
```

Packages
```{r packages}
library(tidyverse)
library(MuMIn)
library(reshape2)
```

Read in t mean model data
```{r data}

dailyT <- read_csv("./InputData/byClaire/allparks_dailyT_2022-11-01.csv")

GSD <- read_csv("./InputData/byClaire/GSD_dailyT_moddata.csv")
RMN <- read_csv("./InputData/byClaire/RMN_dailyT_moddata.csv")
YNP <- read_csv("./InputData/byClaire/YNP_dailyT_moddata.csv")
GNP <- read_csv("./InputData/byClaire/GNP_dailyT_moddata.csv")

```


### GRSD model ###
# All data missing from 2013-09-18 to 2014-08-29

# Fill in climate for GRSD EXCEPT the entire missing year
```{r interpGRSD}

# Get names of all summits in GRSD
summits <- unique(dailyT$summit[dailyT$park=="GSD"]) %>% as.character()
aspects <- c("N","S","E","W")

# This dataframe will be used in the predict function to predict current dependent variable
GSD_nomissyr_df <- filter(GSD,!date%in%seq(as.Date('2013-09-18'),as.Date('2014-08-29'),by='days'))

# This will store observed and predicted data in separate columns for reference
predict_df <- GSD_nomissyr_df %>% select(date,starts_with('tmean'))

# This will store observed data and fill in NAs with predicted data in the same column
filledNAs_df <- predict_df

for(i in summits){
  for(j in aspects){
    
    # Current summit*aspect
    sum_asp <- paste0(i,"_",j)
    
    # Use to remove other summits prism columns
    othersummits <- setdiff(summits,i)
    
    lm_df <- GSD %>% 
      filter(!date%in%seq(as.Date('2013-09-18'),as.Date('2014-08-29'),by='days')) %>% # Only applicable for GRSD
      select(-date) %>% # get rid of date columns
      select(-c(paste0("PR.tmean.",othersummits))) %>% # remove other summits' prism data
      select(-contains("PR.tmin")) %>% 
      select(paste0('tmean.',sum_asp),which(colSums(is.na(.))<10)) %>% # Keep mean t column for current sum*asp and any other columns with fewer than 10 NAs
      na.omit()
    
    # linear model predicting current sum*asp mean temp
    tmp_lm <- lm(formula = paste0("tmean.",i,"_", j,' ~ .'),data = lm_df, na.action = "na.fail")
    tmp_dredge <- dredge(global.model = tmp_lm, extra = "R^2",fixed="doy")
    
    # Select the best model from dredge
    best_model <- get.models(tmp_dredge,subset=1)[[1]]
    
    # Predict t mean for current sum*asp using the best model
    predict_vec <- predict(object = best_model,GSD_nomissyr_df)
    
    # Add predicted data as new column in predict_df and rename
    predict_df <- cbind(predict_df,predict_vec)
    colnames(predict_df)[ncol(predict_df)] <- paste0("pred.tmean.",sum_asp)
    
    # Find column that contains observed t mean data for current sum*asp
    colnum <- which(colnames(filledNAs_df)==paste0('tmean.',sum_asp))
    
    # Fill in NA values in the filledNAs_df
    for(rownum in 1:nrow(filledNAs_df)){
      
       if(is.na(filledNAs_df[rownum,colnum])) {filledNAs_df[rownum,colnum]=predict_vec[rownum]}
      
    }

    print(c(sum_asp,tmp_dredge$`R^2`[1],names(best_model$coefficients)))
    rm(lm_df,tmp_lm,tmp_dredge,best_model,predict_vec,sum_asp,colnum)
    
  } # End aspect loop
} # End summit loop

assign("GSD_predict_df1",predict_df)
assign("GSD_filledNAs_df",filledNAs_df)
sapply(GSD_filledNAs_df, function(x) sum(is.na(x)))

rm(filledNAs_df,predict_df)
```

# Plot predicted data vs observed data for GRSD models that predict all but missing year of data
```{r plotpredictions}

for(i in summits){
  for(j in aspects){
    
    obs <- paste0('tmean.',i,"_",j)
    pred <- paste0('pred.tmean.',i,"_",j)
    print(ggplot(GSD_predict_df1,aes(x=eval(as.name(obs)),y=eval(as.name(pred))))+
      geom_point()+
      geom_abline(slope=1)+
      labs(title=paste(i,j),x=obs,y=pred)+
      theme_bw())
    rm(obs,pred)
  
    }
}
```

# Now fill in the missing year from Great Sand Dunes NP. Since there is not soil T data for this year, these models will rely on PRISM data as independent variables
```{r GRSDmissingyear}

# This will store observed and predicted data in separate columns for reference
predict_df <- GSD %>% select(date,starts_with('tmean'))

GSD_missyr_df <- filter(GSD,date%in%seq(as.Date('2013-09-18'),as.Date('2014-08-29'),by='days'))

GSD_missyr_filledNAs_df <- GSD_missyr_df %>% select(date,starts_with('tmean'))

for(i in summits){
  for(j in aspects){
    
    # Set sum_asp
    sum_asp <- paste0(i,"_",j)
    
    # Create variable of prism column name for sum*asp to be used in model below
    PRcol <- paste0("PR.tmean.",i)
    
    # Create model data.frame
    lm_df <- GSD %>%
      select(paste0("tmean.",sum_asp),starts_with("PR."),doy,doysq) %>% 
      na.omit()
    
    # Create global model
    tmp_lm <- lm(paste0("tmean.",i,"_",j, " ~ . + I(",PRcol,"^2) + I(",PRcol,"^3)"), data = lm_df,  na.action = "na.fail")
    
    # Dredge global model
    tmp_dredge <- dredge(global.model = tmp_lm, extra = "R^2",fixed="doy")
    
    # Pull best model from dredge output
    best_model <- get.models(tmp_dredge,subset=1)[[1]]
    
    # Print stats from best model
    print(c(sum_asp,summary(best_model)$r.squared))
  
    # use the best model to predict temp values for the missing year
    all_predict_vec <- predict(object = best_model,GSD)
    
    # Add predicted data as new column in predict_df and rename
    predict_df <- cbind(predict_df,all_predict_vec)
    colnames(predict_df)[ncol(predict_df)] <- paste0("pred.tmean.",sum_asp)
    
    # Now just predict values for the missing year. Could also subset the above values in predict df 
    missyr_predict_vec <- predict(object=best_model,GSD_missyr_df)
    
    # Find columm for current sum*asp
    colnum <- which(colnames(GSD_missyr_filledNAs_df)==paste0('tmean.',sum_asp))
    
    # Fill in NA values in the filledNAs_df
    for(rownum in 1:nrow(GSD_missyr_filledNAs_df)){
       if(is.na(GSD_missyr_filledNAs_df[rownum,colnum])) {GSD_missyr_filledNAs_df[rownum,colnum]=missyr_predict_vec[rownum]}
      } # End fill NA loop
    
     rm(lm_df,tmp_lm,tmp_dredge,best_model,all_predict_vec,missyr_predict_vec,sum_asp,colnum)
    
  } # end aspect loop
} # end summit loop

# Bind predicted missing year of data to rest of predicted data set
GSD_final_df <- rbind(GSD_filledNAs_df,GSD_missyr_filledNAs_df) %>% arrange(date)
sapply(GSD_final_df, function(x) sum(is.na(x)))

assign("GSD_predict_df2",predict_df)
rm(predict_df)
```

# Plot output from models for missing mean T data for GRSD
```{r plotmissingyearmodels}

for(i in summits){
  for(j in aspects){
    
    obs <- paste0('tmean.',i,"_",j)
    pred <- paste0('pred.tmean.',i,"_",j)
    print(ggplot(GSD_predict_df2,aes(x=eval(as.name(obs)),y=eval(as.name(pred))))+
      geom_point()+
      geom_abline(slope=1)+
      labs(title=paste(i,j),x=obs,y=pred)+
      theme_bw())
    rm(obs,pred)
  }
}

```

#### Model for other parks (not GRSD)
```{r otherparkstmeanmods}

# Get names of all summits in other parks
parks <- c("RMN","YNP","GNP")

for(park in parks){
  summits <- unique(dailyT$summit[dailyT$park==park]) %>% as.character()
  # Get full initial data.frame
  predict_df <- eval(as.name(park)) %>% select(date,starts_with('tmean'))
  filledNAs_df <- predict_df
  
  maxNA <- ifelse(park=="RMN",9,10)

  for(i in summits){
    for(j in aspects){
      
      sum_asp <- paste0(i,"_",j)
      dep_col <- paste0("tmean.",sum_asp)
      PRcol <- paste0("PR.tmean.",i)
      
      # Use to remove other summits prism columns
      othersummits <- setdiff(summits,i)
      
      lm_df <- eval(as.name(park)) %>% 
        select(-c(date)) %>% # get rid of date columns
        select(-c(paste0("PR.tmean.",othersummits))) %>% # remove other summits' prism data
        select(-contains("PR.tmin")) %>% # remove min Temp prism data
        select(paste0('tmean.',sum_asp),which(colSums(is.na(.))<maxNA)) %>% 
        na.omit()

      # GNP has a slightly different model. Use if else to tell loop which model to use based on park
      if(park=="GNP") {
        tmp_lm <- lm(formula = paste0(dep_col,' ~ . + I(',PRcol,'^2) + I(',PRcol,'^3)'),data = lm_df, na.action = "na.fail")
      } else {
        tmp_lm <- lm(formula = paste0("tmean.",i,"_", j,' ~ . '),data = lm_df, na.action = "na.fail")
      }

      # dredge global model
      tmp_dredge <- dredge(global.model = tmp_lm, extra = "R^2",fixed=c(PRcol,"doy"))

      # Get best model from dredge output
      best_model <- get.models(tmp_dredge,subset=1)[[1]]

      # Generate vector of predicted values using the main/original dataframe for park
      predict_vec <- predict(object = best_model,eval(as.name(park)))

      # Bind predicted values to predict DF
      predict_df <- cbind(predict_df,predict_vec)
      colnames(predict_df)[ncol(predict_df)] <- paste0("pred_",sum_asp)

      # Check that output_df has same number of rows (same days) as predict_vec
      # nrow(output_df)==length(predict_vec)
      colnum <- which(colnames(filledNAs_df)==paste0('tmean.',sum_asp))

      # Fill in NA values in the output_df
      for(rownum in 1:nrow(filledNAs_df)){
         if(is.na(filledNAs_df[rownum,colnum])) {filledNAs_df[rownum,colnum]=predict_vec[rownum]}
      }

      print(c(sum_asp,tmp_dredge$`R^2`[1],names(best_model$coefficients)))
      rm(lm_df,tmp_lm,tmp_dredge,best_model,predict_vec,sum_asp,PRcol,colnum,dep_col)

    } # End aspect loop
  } # End summit loop
  assign(paste0(park,"_filledNAs_df"), filledNAs_df)
  assign(paste0(park,"_predict_df"), predict_df)
  rm(filledNAs_df,predict_df,maxNA)
} # end parks loop

```

# Convert model output data.frames from long to wide and write csvs
```{r finalmeanTdfs}

GSD_final <- GSD_final_df %>% 
  melt(id.vars="date") %>% 
  separate(variable,c("summit","aspect"),sep = "_") %>%
  separate(summit,c("x","summit")) %>% 
  select(-x) %>% 
  mutate(park="GSD")
colnames(GSD_final)[4] <- "tmean"
write_csv(GSD_final,"./InputData/byClaire/InterpDailyT/GSD_meandailyT.csv")

RMN_final <- RMN_filledNAs_df %>% 
  melt(id.vars="date") %>% 
  separate(variable,c("summit","aspect"),sep = "_") %>%
  separate(summit,c("x","summit")) %>% 
  select(-x) %>% 
  mutate(park="RMN")
colnames(RMN_final)[4] <- "tmean"
write_csv(RMN_final,"./InputData/byClaire/InterpDailyT/RMN_meandailyT.csv")

YNP_final <- YNP_filledNAs_df %>% 
  melt(id.vars="date") %>% 
  separate(variable,c("summit","aspect"),sep = "_") %>%
  separate(summit,c("x","summit")) %>% 
  select(-x) %>% 
  mutate(park="YNP")
colnames(YNP_final)[4] <- "tmean"
write_csv(YNP_final,"./InputData/byClaire/InterpDailyT/YNP_meandailyT.csv")

GNP_final <- GNP_filledNAs_df %>% 
  melt(id.vars="date") %>% 
  separate(variable,c("summit","aspect"),sep = "_") %>%
  separate(summit,c("x","summit")) %>% 
  select(-x) %>% 
  mutate(park="GNP")
colnames(GNP_final)[4] <- "tmean"
write_csv(GNP_final,"./InputData/byClaire/InterpDailyT/GNP_meandailyT.csv")
```

