
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## This file prepares a dataset that contains total plant cover (summit-aspect-species level), climate (summit-aspect), spatial rang (species-level), and trait data (species-level) for all transitions:
GNP: 2003/2004 -> 2009 -> 2014 -> 2019 (3 transitions)
RMN: 2010 -> 2014 -> 2019 (2 transitions)
YNP: 2011 -> 2016 (1 transition)
GRSD: 2009 -> 2015 -> 2020 (2 transitions)
PECO: 2016 -> 2021 (1 transition)

yr0: Where this is in column names it refers to the growing season period during year of the first census. For example, if a census was completed in 2003, then the yr0.meanT is the mean temperature for the growing season during the 2003 growing season

climyr0: Where this is in column names it refers to the number of snow days leading up to the growing season of the first census. 

# Packages
```{r packages}
rm(list=ls())

library(tidyverse)
library(reshape2)
library(lubridate)
library(ggfortify)
```


# Read-in data
```{r data}
siteinfo = read_csv("./Data/siteInfo/summitcoordinates.csv") %>% 
  dplyr::select(park,summit,sitelat=lat,sitelong=long,siteelev=elev)
parks_ns = c("GNP","YNP","RMN","GSD","PEC")

vascplantcover = read_csv(paste0("Data/IntermediateData/vascplantcover_final.csv"))

# water balance data
wbdata = read_csv("./Data/IntermediateData/dailyWBdata.csv")

# Growing season dates (month-day)
t.gs.start.date = "07-01"
t.grwszn = seq(as.Date(t.gs.start.date,format='%m-%d'),as.Date("09-25",format='%m-%d'),by="day")
t.grwszn = format(t.grwszn,"%m-%d")

ppt.gs.start.date = "06-01"
ppt.grwszn = seq(as.Date(ppt.gs.start.date,format='%m-%d'),as.Date("09-25",format='%m-%d'),by="day")
ppt.grwszn = format(ppt.grwszn,"%m-%d")
```


# Glacier National Park
- 2003/2004 -> 2009 -> 2014 -> 2019
```{r GNP}

gnp_veg <- filter(vascplantcover, park=="GNP")
gnp_clim <- read_csv("./Data/IntermediateData/FilledTempData/GNP_meandailyT.csv") 
summits <- unique(gnp_veg$summit)

gnp_Ts <- NULL

for(s in summits){
  
  out = NULL
  veg_df = filter(gnp_veg,summit==s)
  years = unique(c(veg_df$startyear,veg_df$endyear))
  transitions = length(years)-1
  
  for(i in 1:transitions){
    
    startyr = years[i]
    endyr = years[i + 1]
    
    cover = filter(veg_df,startyear==startyr) 
    
    clim = gnp_clim %>% 
      # Create year and month-day columns
      mutate(year = year(date), md = format(date,"%m-%d")) %>% 
      # filter for growing season dates in current year range
      filter(md%in%t.grwszn & year<=endyr&year>=startyr & summit==s) %>% 
      group_by(year,summit,aspect) %>% 
      # Find mean growing season temperature by year
      dplyr::summarise(tmean=round(mean(tmean,na.rm=T),3),.groups = "drop") %>% 
      ungroup() %>% 
      group_by(summit,aspect) %>% 
      dplyr::summarise(min.tmean=min(tmean), # minimum mean growing season temperature
                       mean.tmean=mean(tmean),
                       max.tmean=max(tmean),
                       .groups = "drop") %>% 
      ungroup()
    
    tmp_clim = merge(cover,clim,by=c("summit","aspect"),all.x = T)
    
    out = rbind(out,tmp_clim)
   
    rm(tmp_clim) 
  }
  
  gnp_Ts = rbind(gnp_Ts,out)
  rm(out)
}

```

# Rocky Mountain National Park
- 2010 -> 2014 -> 2019
```{r RMN}
rmn_veg <- filter(vascplantcover, park=="RMN")
rmn_clim <- read_csv("./Data/IntermediateData/FilledTempData/RMN_meandailyT.csv") 
summits <- unique(rmn_veg$summit)

rmn_Ts <- NULL

for(s in summits){
  
  out = NULL
  veg_df = filter(rmn_veg,summit==s)
  years = unique(c(veg_df$startyear,veg_df$endyear))
  transitions = length(years)-1
  
  for(i in 1:transitions){
    
    startyr = years[i]
    endyr = years[i + 1]
    
    cover = filter(veg_df,startyear==startyr) 
    
    clim = rmn_clim %>% 
      # Create year and month-day columns
      mutate(year = year(date), md = format(date,"%m-%d")) %>% 
      # filter for growing season dates in current year range
      filter(md%in%t.grwszn & year<=endyr&year>=startyr & summit==s) %>% 
      group_by(year,summit,aspect) %>% 
      # Find mean growing season temperature by year
      dplyr::summarise(tmean=round(mean(tmean,na.rm=T),3),.groups = "drop") %>% 
      ungroup() %>% 
      group_by(summit,aspect) %>% 
      dplyr::summarise(min.tmean=min(tmean), # minimum mean growing season temperature
                       mean.tmean=mean(tmean),
                       max.tmean=max(tmean),
                       .groups = "drop") %>% 
      ungroup()
    
    tmp_clim = merge(cover,clim,by=c("summit","aspect"),all.x = T)
    
    out = rbind(out,tmp_clim)
   
    rm(tmp_clim) 
  }
  
  rmn_Ts = rbind(rmn_Ts,out)
  rm(out)
}

```

# Yellowstone National Park
- 2011 -> 2016
```{r YNP}
ynp_veg <- filter(vascplantcover, park=="YNP")
ynp_clim <- read_csv("./Data/IntermediateData/FilledTempData/YNP_meandailyT.csv") 
summits <- unique(ynp_veg$summit)

ynp_Ts <- NULL

for(s in summits){
  
  out = NULL
  veg_df = filter(ynp_veg,summit==s)
  years = unique(c(veg_df$startyear,veg_df$endyear))
  transitions = length(years)-1
  
  for(i in 1:transitions){
    
    startyr = years[i]
    endyr = years[i + 1]
    
    cover = filter(veg_df,startyear==startyr) 
    
    clim = ynp_clim %>% 
      # Create year and month-day columns
      mutate(year = year(date), md = format(date,"%m-%d")) %>% 
      # filter for growing season dates in current year range
      filter(md%in%t.grwszn & year<=endyr&year>=startyr & summit==s) %>% 
      group_by(year,summit,aspect) %>% 
      # Find mean growing season temperature by year
      dplyr::summarise(tmean=round(mean(tmean,na.rm=T),3),.groups = "drop") %>% 
      ungroup() %>% 
      group_by(summit,aspect) %>% 
      dplyr::summarise(min.tmean=min(tmean), # minimum mean growing season temperature
                       mean.tmean=mean(tmean),
                       max.tmean=max(tmean),
                       .groups = "drop") %>% 
      ungroup()
    
    tmp_clim = merge(cover,clim,by=c("summit","aspect"),all.x = T)
    
    out = rbind(out,tmp_clim)
   
    rm(tmp_clim) 
  }
  
  ynp_Ts = rbind(ynp_Ts,out)
  rm(out)
}

```

# Great Sand Dunes National Park
- 2009 -> 2015 -> 2020
```{r GSD}
gsd_veg <- filter(vascplantcover, park=="GSD")
gsd_clim <- read_csv("./Data/IntermediateData/FilledTempData/GSD_meandailyT.csv") 
summits <- unique(gsd_veg$summit)

gsd_Ts <- NULL

for(s in summits){
  
  out = NULL
  veg_df = filter(gsd_veg,summit==s)
  years = unique(c(veg_df$startyear,veg_df$endyear))
  transitions = length(years)-1
  
  for(i in 1:transitions){
    
    startyr = years[i]
    endyr = years[i + 1]
    
    cover = filter(veg_df,startyear==startyr) 
    
    clim = gsd_clim %>% 
      # Create year and month-day columns
      mutate(year = year(date), md = format(date,"%m-%d")) %>% 
      # filter for growing season dates in current year range
      filter(md%in%t.grwszn & year<=endyr&year>=startyr & summit==s) %>% 
      group_by(year,summit,aspect) %>% 
      # Find mean growing season temperature by year
      dplyr::summarise(tmean=round(mean(tmean,na.rm=T),3),.groups = "drop") %>% 
      ungroup() %>% 
      group_by(summit,aspect) %>% 
      dplyr::summarise(min.tmean=min(tmean), # minimum mean growing season temperature
                       mean.tmean=mean(tmean),
                       max.tmean=max(tmean),
                       .groups = "drop") %>% 
      ungroup()
    
    tmp_clim = merge(cover,clim,by=c("summit","aspect"),all.x = T)
    
    out = rbind(out,tmp_clim)
   
    rm(tmp_clim) 
  }
  
  gsd_Ts = rbind(gsd_Ts,out)
  rm(out)
}

```

# Pecos Wilderness area
- 2016 -> 2021
```{r PEC}
pec_veg <- filter(vascplantcover, park=="PEC")

pec_clim <- read_csv("./Data/IntermediateData/allparks_dailyT_final.csv") %>% 
  filter(park=="PEC") %>% 
  mutate(year=year(date), md = as.character(format(date,"%m-%d"))) %>% 
  filter(md%in%t.grwszn) 

summits <- unique(pec_veg$summit)

pec_Ts <- NULL

for(s in summits){
  
  out = NULL
  veg_df = filter(pec_veg,summit==s)
  years = unique(c(veg_df$startyear,veg_df$endyear))
  transitions = length(years)-1
  
  for(i in 1:transitions){
    
    startyr = years[i]
    endyr = years[i + 1]
    
    cover = filter(veg_df,startyear==startyr) 
    
    clim = pec_clim %>% 
      # filter for growing season dates in current year range
      filter(year<=endyr & year>=startyr & summit==s) %>% 
      group_by(year,summit,aspect) %>% 
      # Find mean growing season temperature by year
      dplyr::summarise(tmean=round(mean(tmean,na.rm=T),3),.groups = "drop") %>% 
      ungroup() %>% 
      group_by(summit,aspect) %>% 
      dplyr::summarise(min.tmean=min(tmean), # minimum mean growing season temperature
                       mean.tmean=mean(tmean),
                       max.tmean=max(tmean),
                       .groups = "drop") %>% 
      ungroup()
    
    tmp_clim = merge(cover,clim,by=c("summit","aspect"),all.x = T)
    
    out = rbind(out,tmp_clim)
   
    rm(tmp_clim) 
  }
  
  pec_Ts = rbind(pec_Ts,out)
  rm(out)

  }

```

Merge veg data together for all parks
```{r merge_all_veg}

all_parks <- rbind(gnp_Ts,gsd_Ts,rmn_Ts,ynp_Ts,pec_Ts) %>% 
  merge(.,siteinfo,by=c("park","summit"))

```

## Now read in species-level anaylsis data frame, keep the WB model data and PCA data, bind to this dataframe
```{r WBandPCAcols}
spp_df = read_csv("Data/AnalysisData/specieslevel_analysisdata_final.csv") %>% 
  select(park,summit,aspect,startyear,climPC1,climPC2,contains(c("gdd","snowdays","aet","pet",".gs.",".wy."))) %>% 
  unique()

all_parks_WBdata = merge(all_parks,spp_df,by=c("park","summit","aspect","startyear"))

write_csv(all_parks_WBdata,"Data/AnalysisData/vascplantcover_final.csv")
```



