
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## This file prepares a dataset that contains vegetation (summit-aspect-species level), climate (summit-aspect), spatial range (species-level), and trait data (species-level) for all transitions:
GNP: 2003/2004 -> 2009 -> 2014 -> 2019 (3 transitions)
RMN: 2010 -> 2014 -> 2019 (2 transitions)
YNP: 2011 -> 2016 (1 transition)
GRSD: 2009 -> 2015 -> 2020 (2 transitions)
PECO: 2016 -> 2021 (1 transition)

yr0: Where this is in column names, it refers to the growing season period during year of the first census. For example, if a census was completed in 2003, then the yr0.meanT is the mean temperature for the growing season during the 2003 growing season

climyr0: Where this is in column names, it refers to the number of snow days leading up to census period

h20yr0: Where this is in column names, it refers to total precipitation in the water year leading up to the census period


## load packages
```{r packages}
rm(list=ls())

library(tidyverse)
library(reshape2)
library(lubridate)
library(ggfortify)

# parks north to south
parks_ns = c("GNP","YNP","RMN","GSD","PEC")
```

## Read in data
```{r data}
siteinfo = read_csv("./Data/siteInfo/summitcoordinates.csv") %>% 
  dplyr::select(park,summit,sitelat=lat,sitelong=long,siteelev=elev)

traits = read_csv("./Data/speciestraits/allparks_speclist_20220124 traits.csv")

veg <- read_csv("./Data/IntermediateData/all_1x1veg_w0s_final.csv") %>% 
  group_by(park,summit,aspect,year,sciname) %>% 
  summarise(cover=mean(cover)) %>% 
  merge(.,siteinfo,by=c("park","summit")) %>% 
  merge(.,traits) %>%
  # Clean up data based on what we want going into the models
  filter(!str_detect(sciname," sp$")) %>% 
  filter(tax_unit!='lichen'&habit!="clubmoss"&habit!="fern")

veg$park <- factor(veg$park,levels = parks_ns)

# Range data
ranges = read_csv("./Data/IntermediateData/species_range_summary.csv")
ranges$sciname <- str_to_sentence(ranges$sciname)
ranges = ranges[ranges$n_obs>1,]
ranges = select(ranges,-c(n_obs,n_obs1090))

# water balance data
wbdata = read_csv("./Data/IntermediateData/dailyWBdata.csv")

# Growing season dates (month-day)
# Temperature
t.gs.start.date = "07-01"
t.grwszn = seq(as.Date(t.gs.start.date,format='%m-%d'),as.Date("09-25",format='%m-%d'),by="day")
t.grwszn = format(t.grwszn,"%m-%d")

# Precip
ppt.gs.start.date = "06-01"
ppt.grwszn = seq(as.Date(ppt.gs.start.date,format='%m-%d'),as.Date("09-25",format='%m-%d'),by="day")
ppt.grwszn = format(ppt.grwszn,"%m-%d")
```

# Glacier National Park
- 2003/2004 -> 2009 -> 2014 -> 2019
```{r GNP}

gnp_veg <- filter(veg, park=="GNP")
gnp_clim <- read_csv("./Data/IntermediateData/FilledTempData/GNP_meandailyT.csv") 
summits <- unique(gnp_veg$summit)

gnp_Ts <- NULL

for(s in summits){
  
  # Initiate dataframe
  out = NULL
  
  # Filter GNP data for current summit
  veg_df = filter(gnp_veg,summit==s)
  
  # Get unique years and number of transitions for current summit
  years = unique(veg_df$year) %>% sort()
  transitions = length(years)-1
  
  for(i in 1:transitions){
    
    # first/last year of transition i
    startyr = years[i]
    endyr = years[i + 1]
    
    # Cover in start/end year
    startcover = filter(veg_df,year==startyr) %>% 
      select(sciname,summit,aspect,startyear=year,startcover=cover)
    
    endcover = filter(veg_df,year==endyr) %>% 
      select(sciname,summit,aspect,endyear=year,endcover=cover)
    
    # Merge two above dfs
    tmp = merge(startcover,endcover,by=c("sciname","summit","aspect")) %>% 
      mutate(transition=paste0(s,i))
    
    # get clim data for the current transition period
    clim = gnp_clim %>% 
      # Create year and month-day columns
      mutate(year = year(date), md = format(date,"%m-%d")) %>% 
      # filter for growing season dates in current year range
      filter(md%in%t.grwszn & year<=endyr & year>=startyr & summit==s) %>% 
      group_by(year,summit,aspect) %>% 
      # Find mean growing season temperature by year
      dplyr::summarise(tmean=round(mean(tmean,na.rm=T),3),.groups = "drop") %>% 
      ungroup() %>% 
      # Then find min, max, and mean for growing season values for the whole 5 year period
      group_by(summit,aspect) %>% 
      dplyr::summarise(min.tmean=min(tmean), # minimum mean growing season temperature
                       mean.tmean=mean(tmean),
                       max.tmean=max(tmean),
                       .groups = "drop") %>% 
      ungroup()
    
    # merge with the veg data
    tmp_clim = merge(tmp,clim,by=c("summit","aspect"),all.x = T)
    
    out = rbind(out,tmp_clim)
   
    rm(tmp_clim) 
  }
  
  gnp_Ts = rbind(gnp_Ts,out)
  rm(out)
}
rm(startyr,endyr,years,summits,s,i,startcover,endcover,tmp,clim)
```

# Rocky Mountain National Park
- 2010 -> 2014 -> 2019
```{r RMN}

rmn_veg <- filter(veg, park=="RMN")
summits <- unique(rmn_veg$summit)
rmn_clim <- read_csv("./Data/IntermediateData/FilledTempData/RMN_meandailyT.csv")

rmn_Ts <- NULL

for(s in summits){
  
  out = NULL
  veg_df = filter(rmn_veg,summit==s)
  years = unique(veg_df$year) %>% sort()
  
  for(i in 1:(length(years)-1)){
    
    startyr = years[i]
    endyr = years[i + 1]
    
    startcover = filter(veg_df,year==startyr) %>% 
      select(sciname,summit,aspect,startyear=year,startcover=cover)
    
    endcover = filter(veg_df,year==endyr) %>% 
      select(sciname,summit,aspect,endyear=year,endcover=cover)
    
    tmp = merge(startcover,endcover,by=c("sciname","summit","aspect")) %>% 
      mutate(transition=paste0(s,i))
    
    clim = rmn_clim %>% 
      mutate(year = year(date),
             md = format(date,"%m-%d")) %>% 
      filter(md%in%t.grwszn & year<=endyr&year>=startyr & summit==s) %>% 
      group_by(year,summit,aspect) %>% 
      dplyr::summarise(tmean=round(mean(tmean,na.rm=T),3),.groups = "drop") %>% 
      # ungroup() %>% 
      group_by(summit,aspect) %>% 
      dplyr::summarise(min.tmean=min(tmean),
                       mean.tmean=mean(tmean),
                       max.tmean=max(tmean),
                       .groups = "drop")
    
    tmp_clim = merge(tmp,clim,by=c("summit","aspect"),all.x = T)
    
    out = rbind(out,tmp_clim)
    rm(tmp_clim)
  }
  
  rmn_Ts = rbind(rmn_Ts,out)
  rm(out)
  
}
rm(startyr,endyr,years,summits,s,i,startcover,endcover,tmp,clim)
```

# Yellowstone National Park
- 2011 -> 2016
```{r YNP}

ynp_veg <- filter(veg, park=="YNP")
summits <- unique(ynp_veg$summit)
ynp_clim <- read_csv("./Data/IntermediateData/FilledTempData/YNP_meandailyT.csv")

ynp_Ts <- NULL

for(s in summits){
  
  out = NULL
  veg_df = filter(ynp_veg,summit==s)
  years = unique(veg_df$year) %>% sort()
  
  for(i in 1:(length(years)-1)){
    
    startyr = years[i]
    endyr = years[i + 1]
    
    startcover = filter(veg_df,year==startyr) %>% 
      select(sciname,summit,aspect,startyear=year,startcover=cover)
    
    endcover = filter(veg_df,year==endyr) %>% 
      select(sciname,summit,aspect,endyear=year,endcover=cover)
    
    tmp = merge(startcover,endcover,by=c("sciname","summit","aspect")) %>% 
      mutate(transition=paste0(s,i))
    
    clim = ynp_clim %>% 
      mutate(year = year(date),
             md = format(date,"%m-%d")) %>% 
      filter(md%in%t.grwszn & year<=endyr&year>=startyr & summit==s) %>% 
      group_by(year,summit,aspect) %>% 
      dplyr::summarise(tmean=round(mean(tmean,na.rm=T),3),.groups = "drop") %>% 
      # ungroup() %>% 
      group_by(summit,aspect) %>% 
      dplyr::summarise(min.tmean=min(tmean),
                       mean.tmean=mean(tmean),
                       max.tmean=max(tmean),
                       .groups = "drop")
    
    tmp_clim = merge(tmp,clim,by=c("summit","aspect"),all.x = T)
    
    out = rbind(out,tmp_clim)
    rm(tmp_clim)
  }
  
  ynp_Ts = rbind(ynp_Ts,out)
  rm(out)
  
}
rm(startyr,endyr,years,summits,s,i,startcover,endcover,tmp,clim)
```

## Great Sand Dunes National Park
- 2009 -> 2015 -> 2020
```{r GSD}

gsd_veg <- filter(veg, park=="GSD")
summits <- unique(gsd_veg$summit)
gsd_clim <- read_csv("./Data/IntermediateData/FilledTempData/GSD_meandailyT.csv")

gsd_Ts <- NULL

for(s in summits){
  
  out = NULL
  veg_df = filter(gsd_veg,summit==s)
  years = unique(veg_df$year) %>% sort()
  
  for(i in 1:(length(years)-1)){
    
    startyr = years[i]
    endyr = years[i + 1]
    
    startcover = filter(veg_df,year==startyr) %>% 
      select(sciname,summit,aspect,startyear=year,startcover=cover)
    
    endcover = filter(veg_df,year==endyr) %>% 
      select(sciname,summit,aspect,endyear=year,endcover=cover)
    
    tmp = merge(startcover,endcover,by=c("sciname","summit","aspect")) %>% 
      mutate(transition=paste0(s,i))
    
    clim = gsd_clim %>% 
      mutate(year = year(date),
             md = format(date,"%m-%d")) %>% 
      filter(md%in%t.grwszn & year<=endyr&year>=startyr & summit==s) %>% 
      group_by(year,summit,aspect) %>% 
      dplyr::summarise(tmean=round(mean(tmean,na.rm=T),3),.groups = "drop") %>% 
      # ungroup() %>% 
      group_by(summit,aspect) %>% 
      dplyr::summarise(min.tmean=min(tmean),
                       mean.tmean=mean(tmean),
                       max.tmean=max(tmean),.groups = "drop")
    
    tmp_clim = merge(tmp,clim,by=c("summit","aspect"),all.x = T)
    
    out = rbind(out,tmp_clim)
    rm(tmp_clim)
  }
  
  gsd_Ts = rbind(gsd_Ts,out)
  rm(out)
  
}
rm(startyr,endyr,years,summits,s,i,startcover,endcover,tmp,clim)
```

## Pecos Wilderness area
- 2016 -> 2021
```{r PEC}

pec_veg <- filter(veg, park=="PEC")
summits <- unique(pec_veg$summit)
pec_clim <- read_csv("./Data/climate/NPS_IMD_GLORIA_SoilTemperature_2288176_DataPackage_UPDATED_Pecos/NPS_IMD_GLORIA_SoilTemperature_2288176_Raw-dataset.csv") %>% 
  filter(Park=="USFS_PEC") %>% 
  mutate(park="PEC",
         date=as.Date(DateTime)) %>% 
  select(date,park,summit=Summit,aspect=Plot,tempC=Value) %>% 
  group_by(date,park,summit,aspect) %>% 
  dplyr::summarise(tmean=mean(tempC),.groups = "drop")

pec_Ts <- NULL

for(s in summits){

  out = NULL
  veg_df = filter(pec_veg,summit==s)
  years = unique(veg_df$year) %>% sort()

  for(i in 1:(length(years)-1)){

    startyr = years[i]
    endyr = years[i + 1]

    startcover = filter(veg_df,year==startyr) %>%
      select(sciname,summit,aspect,startyear=year,startcover=cover)

    endcover = filter(veg_df,year==endyr) %>%
      select(sciname,summit,aspect,endyear=year,endcover=cover)

    tmp = merge(startcover,endcover,by=c("sciname","summit","aspect")) %>%
      mutate(transition=paste0(s,i))

    clim = pec_clim %>%
      mutate(year = year(date),
             md = format(date,"%m-%d")) %>%
      filter(md%in%t.grwszn & year<=endyr&year>=startyr & summit==s) %>%
      group_by(year,summit,aspect) %>%
      dplyr::summarise(tmean=round(mean(tmean,na.rm=T),3),.groups = "drop") %>%
      # ungroup() %>%
      group_by(summit,aspect) %>%
      dplyr::summarise(min.tmean=min(tmean),
                       mean.tmean=mean(tmean),
                       max.tmean=max(tmean),
                       .groups = "drop")

    tmp_clim = merge(tmp,clim,by=c("summit","aspect"),all.x = T)

    out = rbind(out,tmp_clim)
    rm(tmp_clim)
  }

  pec_Ts = rbind(pec_Ts,out)
  rm(out)

}

rm(startyr,endyr,years,summits,s,i,startcover,endcover,tmp,clim)
```

## Merge veg data together for all parks
```{r merge_all_veg}

traits <- read_csv("./Data/speciesTraits/allparks_speclist_20220124 traits.csv") 
siteinfo <- read_csv("./Data/IntermediateData/all_1x1veg_no0s_final.csv") %>% 
  select(park,summit,aspect) %>% unique() %>% merge(.,siteinfo)

all_parks <- rbind(gnp_Ts,gsd_Ts,rmn_Ts,ynp_Ts,pec_Ts) %>% 
  filter(!str_detect(sciname," sp$")) %>% 
  merge(.,traits,by="sciname") %>% 
  merge(.,siteinfo,by=c("summit","aspect")) %>% 
  merge(.,ranges,by="sciname",all.x=T) %>%
  mutate(changecover=(endcover-startcover),
         propchangecover=changecover/(startcover+0.1)) %>%
  select(park,summit,aspect,sitelat,siteelev,
         sciname,family,genus,species,
         perenneating=perennating,leaves,habit,taxunit=tax_unit,
         transition,startyear,startcover,endyear,endcover,
         changecover,propchangecover,
         latmin1090,latmax1090,latmean1090,latmedian1090,latrange1090,
         min.tmean,mean.tmean,max.tmean)

```


# Now bring in WB model water/snow year data
```{r snow_wbmdata}

summits <- unique(all_parks$summit)
snowdays <- NULL
h20yppt <- NULL

start.h20y = "10-01"
end.h20y = "09-30"

for(s in summits){
  
  years = unique(veg$year[veg$summit==s]) %>% sort()
  transitions = length(years)-1
  
  for(i in 1:transitions){ # start snow days for transition i loop
    
    startyr = years[i]
    endyr = years[i+1]
      
    # Filter WB data for current transition time frame
    tmp = filter(wbdata,str_detect(ID,s)) %>% # get WB model data for current summit
        separate(ID,into = c("summit","aspect")) %>% # separate ID column into separate summit and aspect coumns
        mutate(year = year(date),
              month=month(date)) %>%
        filter(year>=(startyr-1) & year <= endyr) # year data for the year prior to the census year up to the year of the following census year
      tmp$month = as.numeric(tmp$month) # Turn month into a numeric variable

      ## Fill in water year column -- used as grouping variable in next section
      tmp$h20yr <- NA
      yrs = unique(tmp$year)

      for(yr in 1:(length(yrs)-1)){ # snow years loop
        # start of water year
        start <- as.Date(paste(yrs[yr],start.h20y,sep="-"))

        # end of water uear
        end <- as.Date(paste(yrs[yr+1],end.h20y,sep="-"))

        # Fill in snow year variable for correct data range
        tmp$h20yr[tmp$date>=start&tmp$date<=end]= paste0("snowyr",yr)
      } # end water/snow years loop

      # Remove extra rows of data -- before the start of the initial water year and after the end of that year
      tmp <- tmp %>% filter(!is.na(h20yr))

      snowdaysi <- tmp %>%
        # Add a park column
        mutate(park = unique(veg$park[veg$summit==s])) %>%
        # grop by PSA-water year
        group_by(park,summit,aspect,h20yr) %>%
        # Find the rough number of days with snowcover
        dplyr::summarise(snowdays = sum(PACK>0),.groups = "drop") %>%
        # Group by PSA
        group_by(park,summit,aspect) %>%
        # Find min,max,mean of annual snowcover days within 5-year period
        dplyr::summarise(min.snowdays=min(snowdays),
                         mean.snowdays=mean(snowdays),
                         max.snowdays=max(snowdays),
                         .groups = "drop") %>%
        mutate(transition=paste0(s,i))
      
    # Add to the growing df for summits with multiple transitions
     snowdays <- rbind(snowdays,snowdaysi)
     rm(snowdaysi)
     
     # Repeat for precip
     h20yr.ppti <- tmp %>% 
      mutate(park = unique(veg$park[veg$summit==s])) %>%
      group_by(park,summit,aspect,h20yr) %>%
      dplyr::summarise(wy.ppt = sum(ppt),.groups = "drop") %>%
      group_by(park,summit,aspect) %>%
      dplyr::summarise(min.wy.ppt=min(wy.ppt),
                         mean.wy.ppt=mean(wy.ppt),
                         max.wy.ppt=max(wy.ppt),
                         .groups = "drop") %>%
      mutate(transition=paste0(s,i))
     h20yppt <- rbind(h20yppt,h20yr.ppti)
     rm(h20yr.ppti)
     
  } # end snow days for transition i loop
}
## Add this to the large analysis dataframe
all_parks_h20yr <- merge(all_parks,snowdays,by=c("park","summit","aspect","transition")) %>% 
  merge(.,h20yppt,by=c("park","summit","aspect","transition"))
```


# WB model grw szn precip data data
```{r water_wbmdata}

wbm_grwszn <- NULL

for(s in summits){
  
  years = unique(veg$year[veg$summit==s]) %>% sort()
  transitions = length(years)-1
  
  for(i in 1:transitions){
    
    # Filter WBD for current summit's data
    tmp = filter(wbdata,str_detect(ID,s)) %>%
      # Seperate the summit-aspect into unique columns
      separate(ID,into = c("summit","aspect")) %>%
      # create a month-day column from the dates
      mutate(year = year(date),moday = format(date,"%m-%d")) %>%
      # filter for years of data within transition period
      filter(year>=years[i] & year <= years[i+1])
    
    # soil water availabilty metrics
    tmp_cwd = tmp %>%
      # filter for the precip growing season month-days
      filter(moday%in%ppt.grwszn) %>% 
      # Get annual PET AET, CWD, and PPT growing season values
      group_by(park,summit,aspect,year) %>% 
      dplyr::summarise(pet = mean(PET),
                       aet = mean(AET),
                       cwd = pet-aet,
                       ppt = sum(ppt,na.rm=T),
                       .groups = "drop") %>% 
      group_by(park,summit,aspect) %>% 
      # Then find min, max, and mean of the annual values
      dplyr::summarise(min.pet=min(pet),mean.pet=mean(pet),max.pet=max(pet),
                       min.aet=min(aet),mean.aet=mean(aet),max.aet=max(aet),
                       min.cwd=min(cwd),mean.cwd=mean(cwd),max.cwd=max(cwd),
                       min.gs.ppt=min(ppt),mean.gs.ppt=mean(ppt),max.gs.ppt=max(ppt),
                       .groups = "drop") 
    
    # GDD metrics
    tmp_gdd <- tmp %>% 
      group_by(park,summit,aspect,year) %>% 
      dplyr::summarise(gdd = sum(GDD>0),
                       .groups = "drop") %>% 
      group_by(park,summit,aspect) %>%
      dplyr::summarise(min.gdd=min(gdd),mean.gdd=mean(gdd),max.gdd=max(gdd),.groups = "drop")
    
    tmp_grwszn <- merge(tmp_cwd,tmp_gdd)%>% 
      mutate(transition=paste0(s,i))
  
    wbm_grwszn <- rbind(wbm_grwszn,tmp_grwszn)
    rm(tmp,tmp_cwd,tmp_gdd,tmp_grwszn)    
  }
}

all_parks_h20yr_gs <- merge(all_parks_h20yr,wbm_grwszn,by=c("park","summit","aspect","transition"))

```

# add in PC1 and PC2 based on climate vars
## keep water year and growing season ppt
```{r pc12}
# partial names of climate variables in the dataset
climcols = select(all_parks_h20yr_gs,contains(c(".tmean",".snow",".gdd",".ppt",".pet",".aet"))) %>% 
  names()

pcadf = all_parks_h20yr_gs %>% 
  select(park,summit,aspect,transition,all_of(climcols)) %>% 
  na.omit() %>% 
  unique() %>% 
  mutate(park=as.factor(park))

pcadf$park = factor(pcadf$park,levels=parks_ns)

# Look at corr plot of climate vars
climcor = cor(pcadf[c(climcols)])

# PCA on climate cols of climdf
climpca = prcomp(pcadf[climcols],scale. = T)
  
# Plot PCA
autoplot(climpca,data=pcadf,
         loadings=TRUE,loadings.colour="grey60",
         loadings.label.colour="black",
         loadings.label=T,loadings.label.size=4,
         loadings.label.just=0, colour = 'park',fill="park",frame=TRUE,frame.type="convex",frame.alpha=0.1)+
  theme_classic(base_size = 16)+
  theme(legend.title = element_blank())

pcadf$aspect = factor(climdf$aspect,levels=c("N","S","E","W"))

### Look at PCs 1&2 vs climate vars
pca_df = cbind(pcadf,climpca$x)
plot_cols = names(pca_df[,5:22])


pcdat = as.data.frame(climpca$x) %>% select(climPC1=PC1,climPC2=PC2) %>% 
  cbind(pcadf[,1:4],.)

all_parks_h20yr_gs_pca = merge(all_parks_h20yr_gs,pcdat,by=c("park","summit","aspect","transition"),all.x=T)

write_csv(all_parks_h20yr_gs_pca,paste0("Data/AnalysisData/specieslevel_analysisdata_final.csv"))
```

